Dotted Suffix Trees\\A Structure for Approximate Text Indexing

### 2006-07-04 paper/construction.tex
Merge most recent changes

Constructing the tree is very simple. We start with a suffix tree which includes suffix links and add error trees to it. First, we construct the error tree for the root. This tree is almost a copy of the entire tree, except for two properties:

\begin{enumerate}
\item It does not have the leaf labeled 1 in the original tree. Keeping this leaf would have resulted in a leaf labeled \putstring{.s\$} which does not occur in original string (it is one character too long).
\item For any other leaf $\overline{w\$}$ occuring at position $p$ in the string, we have a new leaf $\overline{.w\$}$ which occurs at position $p-1$ in the string. Therefore, we adjust the labels.

\end{enumerate}

@@

\begin{enumerate}
\item The leaf labeled 1 in the original error tree is not included.
\item Leafs in the copy have a label which is the original label's value minus one.
\item A leaf labeled $p$ is included only if $s_{p-1}$ equals the label of the suffix link.
\end{enumerate}

These conditions are an expression of Lemma~\ref{lemma:suffix-error} and an extension of the conditions for the root. Both are implemented by Algorithm~\ref{algo:copy-subtree}.

\input{copy-subtree-algo}

The only point to note is line~\ref{algoline:copy-subtree:merge}. Since we filter some leafs, without a merging procedure, it would be possible to have nodes which possess only one child. These are removed by merging a child with its parent.

\begin{figure}
\centering
\includegraphics[width=.4\textwidth]{merge}
\caption{Node Merging demonstration}%
\label{fig:merge}
\end{figure}

As discussed above, suffix trees are implemented such that each node is actually just a pair of indices into the original string (exemplified in Figure~\ref{fig:mississippi-nodes}). To merge a child with its parent, we just need to replace both by a node whose starting index is the starting index of the original child minus the length of the parent edge. This process is illustrated in Figure~\ref{fig:merge}.

The construction of the tree using either Ukonnen's or McCreight's algorithm assures that this operation is correct due to the way that the indices are created.

\input{adddotlink-algo}

One can construct the error tree for any node if the tree for the node pointed to by the current node's suffix link has an error tree. This leads to a recursive definition implemented as Algorithm~\ref{algo:addDotLink}.

\input{construct-algo}

Algorithm~~\ref{algo:construct-1} constructs the 1-error dotted tree. The \textit{foreach} loop at line~\ref{algoline:construct-1:foreach} can be implemented using Depth First Search. In fact, the procedure addDotLink only needs to be called for certain nodes (those which do not possess an incoming suffix link). However, there is no readily available way to access only these nodes and the procedure above does little extra work.

\subsubsection*{Time cost}

If the number of nodes in the final tree is $N$ and the alphabet $\Sigma$, then the above algorithm runs in time $\mathcal{O}(N|\Sigma|)$.

The error tree at the root is created in time proportional to the number of nodes it contains. Every other error tree is constructed by looking at an existing one and copying it. Each of these copies need time proportional to the original error tree. Since there are at maximum $|\Sigma|$ incoming suffix links to a node, each error tree is \emph{looked at} a maximum of $|\Sigma|$ times. The sum of all these operations is therefore bounded by $|\Sigma|N$.

@@

\input{copy-subtree-with-dot-link-algo}

The above algorithm can be used to construct trees with any number of errors by iterating it. To construct the (k+1)-error tree from the k-error tree, make an adjusted copy of the tree as above (adjusting leafs and filtering the leafs labeled one) and make this the new root error tree. Then, for every other node, remove the current error tree (it has a level too few). Finally, for every node except the root, construct its error tree as above (adjusting leafs, filtering the leafs labeled one or which do not correspond to positions in the suffix link's label). These error trees will now have the correct strings.

\input{construct-k-algo}

The code for the subtree copy is Algorithm~\ref{algo:copy-subtree-with-dot-link} which differs from Algorithm~\ref{algo:copy-subtree} in that dot link are also followed while copying the tree. The final procedure for constructing k-error trees is Algorithm~\ref{algo:construct-k}.

We note that the method of removing the existing error trees can be improved in practice (the asymptotic bound remains the same) the following way: While making a copy of the tree for construction of the error tree at the root, instead of following a dot link to make a copy, move the error tree to its destination and make the necessary adjustments in place (this move will probably be just an adjustment of pointer depending on the exact implementation). Using this method only leafs labeled one need to be removed (as well as any internal node which becomes unnecessary).

\begin{definition}
We define $N_k$ to be the number of nodes of the k-error dotted tree. $N_0$ is the number of nodes of the suffix tree. If $k$ is known from context we write $N$ for $N_k$.
\end{definition}

>>>>

Starting with a suffix tree which includes suffix links we add error trees to it. First, we construct the error tree for the root which is almost a copy of the entire tree, except for two properties:

\begin{enumerate}
\item It does not have the leaf labeled 1 in the original tree. Keeping this leaf would have resulted in a leaf labeled \putstring{.s\$} which does not occur in original string (it is one character too long).
\item For any other leaf $\overline{w\$}$ occurring at position $p$ in the string, we have a new leaf $\overline{.w\$}$ which occurs at position $p-1$ in the string. Therefore, we adjust the labels.
\end{enumerate}

@@

\begin{enumerate}
\item The leaf labeled 1 in the original error tree is not included.
\item Leafs in the copy have a label which is the original value minus one.
\item A leaf labeled $p$ is included only if $s_{p-1} = a$ ($a$ being the label of the suffix link).
\end{enumerate}

\input{copy-subtree-algo}
\input{adddotlink-algo}%

These conditions are an expression of Lemma~\ref{lemma:suffix-error} and an extension of the conditions for the root. Both are implemented by Algorithm~\ref{algo:copy-subtree}. The only point to note is line~\ref{algoline:copy-subtree:merge}. Since we filter some leaves, without a merging procedure, it would be possible to have nodes which possess only one child. These are removed by merging a child with its parent.
\begin{figure}
\centering
\includegraphics[width=.4\textwidth]{merge}
\caption{Node Merging example}%
\label{fig:merge}
\end{figure}

As discussed above, suffix trees are implemented in a way such that each node stores the start and end indices of the substring on its incoming edge. To merge a child with its parent, we just need to replace both by a node whose starting index is the starting index of the original child minus the length of the parent edge. The construction of the tree using either Ukonnen's or McCreight's algorithm assures that this operation is correct due to the way that the indices are created.

One can construct the error tree for any node if the tree for the node pointed to by the current node's suffix link has an error tree. This leads to a recursive definition implemented as Algorithm~\ref{algo:addDotLink}. This procedure needs to be called for all nodes, which can be done using a simple depth first search on the tree.

\subsubsection*{Time Cost}

Copying a tree takes time proportional to the number of nodes it contains. The error tree at the root is a straightforward copy of the whole tree and every other tree is a copy of an existing one. Since each node can have at most $|\Sigma|$ incoming suffix links, each error tree is \emph{looked at\/} to be copied at most $|\Sigma|$ times. The sum of all these operations is therefore bounded by $|\Sigma|N$. Therefore, if the number of nodes in the final tree is $N$, then the above algorithm runs in time $\mathcal{O}(N|\Sigma|)$.

@@

\input{construct-k-algo}
The above algorithm can be used to construct trees with any number of errors by iterating it. To construct the $(k+1)$-error tree from the $k$-error tree, make an adjusted copy of the tree as above (adjusting leaves and filtering the leaves with label~1) and make this the new root error tree. Then, for every other node, remove the current error tree (it has a level too few). Finally, for every node except the root, construct its error tree as above (adjusting leaves, filtering the leaves with label~1 or which do not correspond to positions in the suffix link's label). These error trees will now have the correct strings. The final procedure for constructing k-error trees is Algorithm~\ref{algo:construct-k}.

We note that the method of removing the existing error trees can be improved in practice (the asymptotic bound remains the same) the following way: While making a copy of the tree for construction of the error tree at the root, instead of following a dot link to make a copy, move the error tree to its destination and make the necessary adjustments in place.% (this move will probably be just an adjustment of pointer depending on the exact implementation). Using this method only leaves labeled one need to be removed (as well as any internal node which becomes unnecessary).

\begin{definition}
We define $N_k$ to be the number of nodes of the $k$-error dotted tree. $N_0$ is the number of nodes of the suffix tree. We will use $N$ for $N_k$ if $k$ is known from context.
\end{definition}

### 2006-07-04 paper/experimental.tex
Count steps
Merge most recent changes

The code was implemented in the C~Language, starting from an open-source implementation of suffix trees~\cite{ansicimplementation}. Three data sets were used:

\begin{enumerate}
\item The English translation of ``Democracy in America'' by Alexis de Tocqueville as natural language example
\item The DNA of S. Cerevesiae (first chromosome)
\item Randomly generated text by a uniform distribution
\end{enumerate}

The machine used for testing was an Intel Pentium~IV running at~2.40GHz with~4GB of main memory.

@@

\begin{figure*}%
\centering
\includegraphics[width=.7\textwidth]{ratio-in=english-01-12.eps}%
\caption{Ratio of number of nodes in k-error and (k+1)-error dotted trees}\label{fig:ratios}%
\end{figure*}

To experimentally verify the average case prediction, we show in Figures~\ref{fig:ratios} the ratios between the k-error and the (k+1)-error dotted trees, for English text (the other datasets show similar behaviour). We can easily see that the experimental values do resemble a logarithm as predicted.

\subsection{Search Time}

For searching, the texts were first indexed and then online searches were performed on top of the structure. To generate input patterns, randomly chosen snippets of the text were taken. These patterns were then altered by substituting a few characters at random. To generate mismatches, one character too many was altered\,\footnote{Since the characters used were not present in the original text, there is no risk of generating a string which \emph{does appear} in the text.}. The search algorithm performed an early exit, without reporting of occurrences (ie, it only reported whether the string exists in the text). Therefore, the number of occurrences had no influence on the search time.

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{search-vary-N.in=all.eps}
\caption{Search time versus text size}\label{fig:search-vary-N}
\end{figure*}

%Figure~\ref{fig:search-vary-m} shows the search time for growing pattern lenghts. It shows the case where the pattern exists and where it does not (an extra error is added). The fact that the search is done using an early-exit strategy explains why non existing patterns take longer than existing patterns. The value presented is the number of character comparisons the algorithm makes.

Figure~\ref{fig:search-vary-N} shows searching for a 15~character long sequences on a 2-error dotted tree. Presented are the number of character comparisons performed (averaged over 10\,000 such searches) while varying the text size. We see that after an initial small growth which is explained by the increasing density of the tree, the search time is roughly constant.

>>>>

The code was implemented in the \CC~Language~\footnote{Code is available upon request.}. Three data sets were used: a set of English texts; the DNA of S. Cerevesiae; and randomly generated text (uniform distribution). For testing we used an Intel Pentium~IV running at~2.40GHz with~4GB of main memory.

@@

\begin{figure*}%
\centering
\subfigure[DNA]{\includegraphics[height=3cm,width=.45\textwidth]{ratio-in=dna-01-12.eps}}%
\subfigure[English]{\includegraphics[height=3cm,width=.45\textwidth]{ratio-in=english-01-12.eps}}\\%
\subfigure[Random]{\includegraphics[height=3cm,width=.45\textwidth]{ratio-in=random-01-12.eps}}%
\caption{Ratio of number of nodes in k-error and (k+1)-error dotted trees}\label{fig:ratios}%
\end{figure*}
%
To experimentally verify the average case prediction, we show in Figures~\ref{fig:ratios} the ratios between the $k$-error and the $(k+1)$-error dotted trees, for all datasets. We can easily see that the experimental values do resemble a logarithm as predicted.
%
\subsection{Search Time}
%
\begin{figure*}
\centering
\subfigure[k=1]{\includegraphics[width=.75\textwidth,height=5cm]{search-vary-N.in=all.k=1.steps.eps}}\\
\subfigure[k=2]{\includegraphics[width=.75\textwidth,height=5cm]{search-vary-N.in=all.k=2.steps.eps}\hspace{-1cm}}
\caption{Search time versus text size}\label{fig:search-vary-N}
\end{figure*}
%
For searching, the texts were first indexed and then searches were performed on top of the structure. The search algorithm performed an early exit, without reporting of occurrences (ie, it only reported whether the string exists in the text). Therefore, the number of occurrences had no influence on the search time.
%
%Figure~\ref{fig:search-vary-m} shows the search time for growing pattern lengths. It shows the case where the pattern exists and where it does not (an extra error is added). The fact that the search is done using an early-exit strategy explains why non existing patterns take longer than existing patterns. The value presented is the number of character comparisons the algorithm makes.

Figure~\ref{fig:search-vary-N} shows searching for a 15~character long sequences on a $k$-error dotted tree. Presented are the number of steps in the algorithm (averaged over 10\,000 such searches) while varying the text size. We see that after an initial small growth which is explained by the increasing density of the tree, the search time is roughly constant.

\subsection{Varying Alphabet Size}
%
\begin{figure*}
\centering
\subfigure[Nodes]{\includegraphics[width=.35\textwidth]{nodes-vary-E.eps}}
%\subfigure[Time]{\includegraphics[width=.25\textwidth]{construction-time-vary-E.eps}}
\subfigure[Ratio]{\label{subfig:nodes-vary-E-ratio}\includegraphics[width=.35\textwidth]{construction-time-ratio-vary-E.eps}}
\caption{Construction time versus $|\Sigma|$}\label{fig:nodes-vary-E}
\end{figure*}%
%
\begin{figure*}
\centering
\subfigure[Time]{\includegraphics[width=.35\textwidth]{search-time-vary-E.eps}}%
\subfigure[Steps]{\includegraphics[width=.35\textwidth]{search-steps-vary-E.eps}}
\caption{Searching versus $|\Sigma|$}\label{fig:search-vary-E}
\end{figure*}%
%
We generated 100\,KB long strings using a uniform distribution. We varied the alphabet from binary through 100 and constructed a~2-error tree. As above, searches were for 15~character long strings, averaged over 10\,000 repetitions.

As predicted, the time to construct one node in the tree (shown in Figure~\ref{subfig:nodes-vary-E-ratio}) grows in a roughly linear fashion. The search time (shown in Figure~\ref{fig:search-vary-E}) does not grow significantly with alphabet size.

### 2006-07-04 paper/introduction.tex
Merge most recent changes

Since their introduction~\cite{DBLP:conf/focs/Weiner73}, suffix trees have been one of the methods of choice for text indexing. Suffix trees can be used solve the following problem: given a text which has been preprocessed, and a pattern, efficiently find the places in the text where this pattern occurs.  However, for many real-life problems this is too restrictive. In this class of problems, one is interested in finding places in the text where an approximate form of the pattern occurs. Several algorithms have been proposed to solve this problem. Navarro presents a survey of the problem and existing approaches to solving it at the time that work was written~\cite{navarro:guided}. Recently there have been some further work in the field, notably the work of M.~Maa\ss{}~\cite{maass:text} which also contains an updated overview of methods for solving the problem in addition to presenting one itself.

@@

>>>>

Since their introduction~\cite{DBLP:conf/focs/Weiner73}, suffix trees have been one of the methods of choice for text indexing. Suffix trees can be used to solve the following problem: given a text which has been preprocessed, and a pattern, efficiently find the places in the text where this pattern occurs.  However, for many real-life problems this is too restrictive. In this class of problems, one is interested in finding places in the text where an approximate form of the pattern occurs. Several algorithms have been proposed to solve this problem. Navarro {\it et al} present a survey of the problem and existing approaches to solving it at the time that work was written~\cite{navarro:guided}.

@@

Amongst competing solutions for this problem, the most relevant is the
work of M.~Maa\ss{}~\cite{maass:text}. His solution takes space
$\bigO(|\Sigma|^k n\log^k n)$ (average case)~\footnote{Maa{\ss}
considers $|\Sigma|$ and $k$ as constant and presents $\bigO(n\log^k
n)$ as a complexity result. However, such analysis ignores the
potentially large impact of alphabet size.} for a search time
$\bigO(m)$.

This work is also related to the work of Cole {\it el
al}~\cite{cole:dictionary} whose initial intuitions resemble
ours. However, they make different time and space tradeoffs to achieve
$\bigO(m+\log\log n+\mathit{occ})$ searching (Hamming distance) with
$\bigO(n\frac{log^k n}{k!})$ space. They also propose an algorithm for
searching using edit distance at a time cost of $\bigO(m+\frac{\log^k
n \log\log n}{k!} +3^k \mathit{occ})$. We improve on that result,
especially for edit distance, by totally removing the dependency
on~$n$ while searching.

Very recently, Navarro and Chavez have presented a completely new
approach based on metric spaces~\cite{navarro:metric}. This approach
achieves, on average, $\mathcal{O}(m^{1+\epsilon})$ search time. Their
structure takes however space $\bigO(m^{1+\sqrt{2}+\epsilon}n)$. In
order to obtain such good bounds it also requires $m$ to be known at
construction time. They have presented a variation which trades time
for space to achieve $\bigO(n)$ space, but at a cost in construction
and search time.

### 2006-07-04 paper/search.tex
Merge most recent changes

To search a string with at maximum k errors in a k-error dotted tree, we descend the tree. Whenever we find a node, we follow both the dot link and the edge which matches the current position of the string we are searching for. If we find a mismatch in the middle of an edge, we continue, decreasing the amount of errors permitted. When presenting the algorithm, we take the liberty of using $s+j$ where $s$ is a string and $j$ an integer to mean the suffix of $s$ starting at position $j$ (as happens in the C~Programming Language).

\input{search-algo}

The process just described is Algorithm~\ref{algo:search}. Reporting of the leafs below a certain node can be done by a Depth First Search. If one is interested in knowing only whether the string occurs, then the lines which report the matches can be substituted by early exits returning true.

This algorithm runs in time $\mathcal{O}(m^{k+1})$: There are at most $\sum_{i=0}{k}{m \choose i}=\mathcal{O}(m^k)$ different paths which must be followed. Following one path in the tree takes $\mathcal{O}(m)$ time. Therefore, the total time is $\mathcal{O}(m^{k+1})$ without the reporting of occurrences (which can be done in time linear to the number of occurrences).

>>>>

Given a pattern to search for, we descend the tree point by point, following the pattern character by character. We represent this by keeping a node and an offset from the start of its incoming link. In an edge, we consider that there is an implicit dot link which goes forward one character. At each point, we can take four possible actions:

\begin{description}
\item[match] Where we descend the tree according to the pattern (may not be possible).
\item[substitution] We follow the dot link (possibly implicit), moving in the pattern.
\item[insertion] We follow the (possibly implicit) dot link, while not moving in the pattern.
\item[deletion] We advance in the pattern, while not moving in the tree.
\end{description}

\input{search-algo-edit}
%
Going down the tree can therefore be seen as a sequence of edition and matching operations: For example, for the string \putstring{ixsi}: \operation{match(i)}, \operation{substitution(x)}, \operation{match(s)}, \operation{match(i)}, where there are at most $k$ elements which are not \operation{match(c)}. There are at most $\sum_{i=1}^k{m \choose i}=\mathcal{O}(m^k)$ ways to combine $k$ edit operations into a string of size $m$. Since there are $3$ different operations (substitution, insertion, and deletion), we have at most $\mathcal{O}(3^k m^k)$ different edition sequences. Each sequence has at most $m+k=\bigO(m)$ (given that $k<m$) elements and therefore the total time to find matches is $\mathcal{O}(3^k m^{k+1})$. Once a match has been found in the tree, a Depth First Search can be used to report all the leaves below the node. This takes time proportional to the number of leaves, ie.\ proportional to the number of edit scripts which can be used to match the pattern to a substring of the text (which can be greater than the number of occurences~\footnote{As often happens, strings of the form $a^m$ serve as examples of pathological behaviour as they can match any position of a string of form $a^n$ in a large number of ways.}). The total search time is therefore $\bigO(3^k m^{k+1} + \mbox{\it ed})$.

The code needs to distinguish whether we are at the end of a branch (ie.\ at a node) or in the middle of one. When presenting the algorithm, we take the liberty of using $s+j$ where $s$ is a string and $j$ an integer to mean the suffix of $s$ starting at position $j$ (as happens in the C~Programming Language). The process just described is Algorithm~\ref{algo:search}.

### 2006-07-04 paper/space.tex
Merge most recent changes

Suffix trees have a number of nodes proportional to the size of the string (which we will refer to as $n$). Dotted trees have more nodes. The worst case is a string of the form $a^n = aaa\dots$ which generates a one-dotted tree taking $\mathcal{O}(n^2)$ nodes (illustrated in Figure~\ref{fig:aaaa}). For a larger number of errors $k$, this string leads to $\mathcal{O}(n^{k+1})$ nodes. We can do better, however, by considering the expected case.

\begin{definition}[Maximum Path Lenght]
$l$ is the length of the deepest branching node in the tree

\[ l = \max_{\overline{w} \in \mathcal{T}} |w| \mbox{ and $\overline{w}$ isn't a leaf} \]
\end{definition}

\begin{definition}[Suffix-Depth]
For any node $\overline{w}$ its suffix-depth is the number of nodes which must be followed to reach the root, which is $|w|$. It is obvious that $l$ is a maximum on the suffix-depth of nodes.
\end{definition}

We now show $N_k = \mathcal{O}(nl^k)$ using induction. It is a known fact that $N_0 = \mathcal{O}(n)$. The algorithm for turning a k-error into a (k+1)-error dotted tree, can be looked at the following way~\footnote{We are here considering that the second \textit{foreach} loop of Algorithm~\ref{algo:construct-k} runs in a specific order which is in fact difficult to code for. However, as an analysis tool, it is a valid assumption.}: First it makes an almost exact copy of the k-error dotted tree and as the error tree at the root and it clears all the other error trees. Then it proceeds in stages, making a (possibly incomplete) copy of this tree spread amongst the nodes at suffix-depth 1. It proceeds to the other nodes in increasing suffix-depths up to $l$. At each such step, the number of nodes is increased by a maximum of $N_k$. Therefore, we start with $N_k$ nodes, make an almost full copy, and copy that at most $l$ times, $N_{k+1}=\mathcal{O}(N_k(l+1))$. Assuming $N_k=\mathcal{O}(nl^k)$ by induction we conclude $N_{k+1}=\mathcal{O}(nl^{k+1})$.

So far, we have achieved little since in the worst case $l=n-1$ as shown above. However, the expected case is $l=\mathcal{O}(\log n)$~\cite{apostolico92selfalignments,szpankowski:unexpected} and we have $N_k=\mathcal{O}(n\log^k n)$.

>>>>

Suffix trees have a number of nodes proportional to the size of the string (which we will refer to as $n$). Dotted trees have more nodes. The worst case is a string of the form $a^n = aaa\dots$ which generates a one-dotted tree taking $\mathcal{O}(n^2)$ nodes (illustrated in Figure~\ref{fig:aaaa}). For a larger number of errors $k$, this string leads to $\mathcal{O}(n^{k+1})$ nodes. An analysis of the average case, however, leads to a much better result.

\begin{definition}[Maximum Path Length, Suffix Depth]
$l$ is the string depth of the deepest branching node in the tree.
%
%\[ l = \max_{\overline{w} \in \mathcal{T}} |w| \mbox{ and $\overline{w}$ isn't a leaf} \]
%\end{definition}

%\begin{definition}[Suffix-Depth]
For any node $\overline{w}$ its suffix-depth is the number of nodes which must be followed to reach the root, which is $|w|$. It is obvious that $l$ is a maximum on the suffix-depth of nodes.
\end{definition}

We now show $N_k = \mathcal{O}(nl^k)$ using induction. It is a known fact that $N_0 = \mathcal{O}(n)$. The algorithm for turning a $k$-error into a $(k+1)$-error dotted tree, can be looked at the following way\,\footnote{We are here considering that the second \textit{foreach} loop of Algorithm~\ref{algo:construct-k} runs in a specific order which is in fact difficult to code for. However, as an analysis tool, it is a valid assumption.}: First it makes an almost exact copy of the k-error dotted tree as the error tree at the root and it clears all the other error trees. Then it proceeds in stages, making a (possibly incomplete) copy of this tree spread amongst the nodes at suffix-depth 1. It proceeds to the other nodes in increasing suffix-depths up to $l$. At each such step, the number of nodes is increased by a maximum of $N_k$. Therefore, we start with $N_k$ nodes, make an almost full copy, and copy that at most $l$ times, $N_{k+1}=\mathcal{O}(N_k(l+1))$. Assuming $N_k=\mathcal{O}(nl^k)$ by induction we conclude $N_{k+1}=\mathcal{O}(nl^{k+1})$.

So far, we have achieved little since in the worst case $l=n-1$ ($s=a^n$). However, the expected case is $l=\mathcal{O}(\log n)$~\cite{apostolico92selfalignments,szpankowski:unexpected} and we have $N_k=\mathcal{O}(n\log^k n)$.

### 2006-07-04 paper/structure.tex
Merge most recent changes

\subsection{Intuition}

\begin{figure}
\centering
\includegraphics[width=.45\textwidth]{figures/mississippi-0}
\caption{Suffix tree representation for the string \putstring{mississippi}}%
\label{fig:mississippi-0}
\end{figure}

Suffix Trees have become a well known structure since their introduction in 1973~\cite{DBLP:conf/focs/Weiner73}. The example in Figure~\ref{fig:mississippi-0} shows the suffix tree for the string \putstring{mississippi}.

@@

For denoting characters we shall use letters from the beginning of the roman alphabet ($a$, $b$, $c$,\ldots) and for strings, we shall use letters from the end of the alphabet ($w$, $z$, \ldots). In what follows we assume that there are two symbols (\$ and $.$) which are not part of $\Sigma$.

\begin{definition}[Concatenation]
$wx$ will denote the string formed by all the characters of $w$ followed by all the characters of $x$ (the usual concatenation operation). $aw$ ($wa$) shall denote the string formed by the character $a$ before (after) the characters of $w$.

@@

\begin{definition}[Patricia tree]
$T$ is a Patricia tree if $T$ is a rooted tree with edge labels from $\Sigma^+$. For each $a \in \Sigma$ and every node $n$ in $T$, there is at most one edge leaving $n$ whose label is $aw$.

Each node in a Patricia tree has a path leading to it which forms a string. If the node $n$ has the leading path $w$, we shall also refer to $n$ as $\overline{w}$.
\end{definition}

%The name Patricia is sometimes spelled PATRICIA, since it was formed from the initial of Practical Algorithm to Retrieve Information Coded in Alphanumeric~\cite{DBLP:journals/jacm/Morrison68}.
%

In this paper, we are interested in the substitution distance (aka Hamming distance):

\begin{definition}[Aproximate Match]
We say that the string $s$ matches the string $t$ at position $p$ with $k$ errors if we can make $k$ substitutions in $s$ to obtain $s'$ which is a substring of $t$ at position $p$.
\end{definition}

\begin{definition}[Pattern]
A pattern is a string over the extended alphabet $\Sigma\cup\{.\}$. We say that a pattern $s$ matches a string $t$ at position $p$, if for each dot in the pattern we can replace it by a character in the original alphabet to obtain $s'$ which is a substring of $t$ at position $p$.
\end{definition}

\begin{definition}[Suffix Tree]
A \emph{suffix tree} for a string $S$ is a Patricia tree whose leaf nodes (those without children) have paths corresponding to all suffixes of the string $S\$ $.
\end{definition}

\begin{definition}[Suffix Link]
A suffix link in a suffix tree is a link from the node $\overline{aw}$ to the node $\overline{w}$. This node has the label $a$.
\end{definition}

In a suffix tree, it is possible to define a suffix link for each internal node. Both Ukkonen's~\cite{ukkonen92constructing} and McCreight's~\cite{mccreight} algorithms are linear time algorithms for constructing a suffix tree which includes suffix links. The suffix links are shown in Figure~\ref{fig:mississippi-0} as dashed arrows with the label in italic.

\begin{figure*}
\includegraphics[width=\textwidth]{figures/mississippi-nodes}
\caption{Suffix Tree for the string \putstring{mississippi}}%
\label{fig:mississippi-nodes}
\end{figure*}

Suffix trees are normally implemented so that each node consists of only two indices into the original string which are the start and end position of its incoming edge. Figure~\ref{fig:mississippi-nodes} illustrates this.

\begin{definition}[Occurence Set]
Given a node $\overline{w}$ in a suffix tree, we call its \emph{occurrence set} the set of indexes in the original string where the string $w$ occurs.
\end{definition}

\begin{definition}[Position Set]
Given a node $\overline{w}$ in a suffix tree, its \emph{position set} is the set formed by taking its occurrence set and adding the length of $w$ to each element.
\end{definition}

@@

\begin{definition}[Error Tree]
For any node $w$, its error tree is the Patricia tree formed by taking its position set, adding one to each element and forming the Patricia tree of the suffixes of the suffixes starting at those positions. If the position set includes the end of the string, that element is removed.

The leafs are labeled by the position of the string in which their paths occur minus $|w| + 1$.
\end{definition}

@@

\begin{definition}[1-error dotted Tree]
We define a \emph{1-error dotted tree} as the tree which is formed by adding to each node in a suffix tree, a new edge labeled by \putstring{.} which points to its error tree. The edge labeled \putstring{.} shall be called a \emph{dot link}.
\end{definition}

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{figures/mississippi-1.eps}
\caption{1-error dotted tree for \putstring{mississippi}}%
\label{fig:mississippi-1}
\end{figure*}

The 1-error dotted tree for \putstring{mississippi} is shown in Figure~\ref{fig:mississippi-1}. The nodes are connected to their error trees by thick diagonal links.

The paths in the dotted tree are paths in the extended alphabet $\Sigma\cup\{.,\$\}$. The notions of \emph{occurrence set}, \emph{position set} and \emph{error tree} are valid for all nodes in a dotted tree considering that a node path may contain dot elements which match any other character.

\begin{definition}[k-error dotted tree]
We define a k-error dotted tree as the tree obtained by adding error trees to each node in the (k-1)-error dotted tree which does not already contain one.
\end{definition}

>>>>

\begin{figure*}
\centering
\includegraphics[width=.6\textwidth]{figures/mississippi-0}
\caption{Suffix tree representation for the string \putstring{mississippi}}%
\label{fig:mississippi-0}
\end{figure*}
\subsection{Intuition}
Suffix Trees have become a well known structure since their introduction in 1973~\cite{DBLP:conf/focs/Weiner73}. The example in Figure~\ref{fig:mississippi-0} shows the suffix tree for the string \putstring{mississippi}.

@@

For denoting characters we shall use letters from the beginning of the roman alphabet ($a$, $b$, $c$,\ldots) and for strings, we shall use letters from the end of the alphabet ($w$, $z$, \ldots). In what follows we assume that there are two special symbols (\$ and $.$) which are not part of $\Sigma$.

\begin{definition}[Concatenation]
$wx$ or $aw$ will denote the usual concatenation operation.

@@

\begin{definition}[Patricia tree]
$T$ is a Patricia tree if $T$ is a rooted tree with edge labels from $\Sigma^+$. For each $a \in \Sigma$ and every node $n$ in $T$, there is at most one edge leaving $n$ whose label starts with $a$. Each node in a Patricia tree has a path leading to it which forms a string. If the node $n$ has the leading path $w$, we shall also refer to $n$ as $\overline{w}$.
\end{definition}

%The name Patricia is sometimes spelled PATRICIA, since it was formed from the initial of Practical Algorithm to Retrieve Information Coded in Alphanumeric~\cite{DBLP:journals/jacm/Morrison68}.
%
In this work we are interested in approximate matching using edition (or Levenshtein) distance:

\begin{definition}[Aproximate Match]
We say that the string $s$ matches the string $t$ at position $p$ with $k$ errors if we can make $k$ modifications in $s$ to obtain $s'$ which is a substring of $t$ at position $p$. A modification is either deletion, insertion or substitution of one character.
\end{definition}

\begin{definition}[Suffix Tree, Suffix Link]
A \emph{suffix tree} for a string $S$ is a Patricia tree whose leaf nodes (those without children) have paths corresponding to all suffixes of the string $S\$ $. A suffix link in a suffix tree is a link from the node $\overline{aw}$ to the node $\overline{w}$. This link has the label $a$.
\end{definition}

In a suffix tree, all internal nodes have a well defined suffix link. Both Ukkonen's~\cite{ukkonen92constructing} and McCreight's~\cite{mccreight} algorithms are linear time algorithms for constructing a suffix tree which includes suffix links. The suffix links are shown in Figure~\ref{fig:mississippi-0} as dashed arrows with the label in italic.

Suffix trees are normally implemented by associating with each node two indices into the original string which are the start and end position of its incoming edge.

\begin{definition}[Point in a suffix tree]
A \emph{point} in a suffix tree is either a node in the tree or a point in the middle of an edge.
\end{definition}

\begin{definition}[Occurrence Set, Position Set]
Given a node $\overline{w}$ in a suffix tree, we call its \emph{occurrence set} the set of indexes in the original string where the string $w$ occurs.

Given a node $\overline{w}$ in a suffix tree, its \emph{position set} is the set formed by taking its occurrence set and adding the length of $w$ to each element.
\end{definition}

@@

\begin{definition}[Error Tree]
For any node $w$, its error tree is the Patricia tree formed by taking its position set, adding one to each element and forming the Patricia tree of the suffixes starting at those positions. If the position set includes the end of the string, that element is removed.

The leaves are labeled by the position of the string in which their paths occur minus $|w| + 1$.
\end{definition}

@@

\begin{definition}[1-error dotted Tree]
We define a \emph{1-error dotted tree} as the tree which is formed by adding to each node in a suffix tree, a new edge labeled by \putstring{$\cdot$} which points to its error tree. The edge labeled \putstring{$\cdot$} shall be called a \emph{dot link}.
\end{definition}

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{figures/mississippi-1.eps}%
\caption{1-error dotted tree for \putstring{mississippi}}%
\label{fig:mississippi-1}%
\end{figure*}%
The 1-error dotted tree for \putstring{mississippi} is shown in Figure~\ref{fig:mississippi-1}. The nodes are connected to their error trees by thick diagonal links.

The paths in the dotted tree are paths in the extended alphabet $\Sigma\cup\{.,\$\}$. The notions of \emph{occurrence set}, \emph{position set} and \emph{error tree} are valid for all nodes in a dotted tree.

\begin{definition}[k-error dotted tree]
We define a $k$-error dotted tree as the tree obtained by adding error trees to each node in the $(k-1)$-error dotted tree which does not already contain one.
\end{definition}
